---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Setting programming environment
- Clean up
```{r}
rm(list=ls())
```


- Load libraries
```{r}
library(lubridate)
library(dplyr)
```



## Loading and preprocessing the data
- Read raw dataset
```{r}
activity_raw <- read.csv("activity.csv")
```


- Convert strings to lubridates
```{r}
activity <- activity_raw %>% mutate(date=ymd(date))
```



## What is mean total number of steps taken per day?
- Calculate the total number of steps taken per day
```{r}
steps_per_day <- activity %>% 
        group_by(date) %>% 
        do (
                mutate(., sum=sum(steps, na.rm = T)
                       , mean=mean(steps, na.rm = T)
                       , median=median(steps, na.rm = T)
                )
        ) %>%
        distinct(date)

steps_per_day %>% select(date, sum)
```


- Make a histogram of the total number of steps taken each day
```{r}
total_days <- as.numeric(max(steps_per_day$date) - min(steps_per_day$date) + 1)
hist(steps_per_day$sum
        , main = paste("Total number of steps per day","\n(",total_days,"-day period)")
        , xlab="Steps per day"
        , ylab="Frequency of days"
)
```


- Calculate and report the mean and median of the total number of steps taken per day
```{r}
steps_per_day %>% select(date, mean, median)
```



## What is the average daily activity pattern?
- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
five_m_int <- activity %>% 
        group_by(interval) %>% 
        mutate( mean=mean(steps, na.rm = T)) %>%
        distinct(interval)

with(five_m_int, 
        plot(interval, mean, type = "l"
                , main="Steps per 5-minute interval"
                , xlab="Intervals"
                , ylab="Average number of steps"
        )
)
```


- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mxmean <- max(five_m_int$mean)
five_m_int[ five_m_int$mean == mxmean, c("interval", "mean")]
```



## Imputing missing values
- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
n_complete <- sum(complete.cases(activity))
n_total <- dim(activity)[1]
n_total - n_complete
```


- Devise a strategy for filling in all of the missing values in the dataset. 

The strategy for imputing the missing values in the column *steps* will be to replace all NAs by the mean number of steps for that 5-minute interval (averaged across all days).


- Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
imputation_values <- five_m_int  %>% select(interval, mean)
activity_new <- left_join(
                activity
                , imputation_values
                , by="interval"
                , copy=T
        )

index <- which(is.na(activity_new$steps))
nmax <- length(index)
for (i in 1:nmax){
        index_i <- index[i]
        activity_new$steps[index_i] <- activity_new$mean[index_i] 
}

activity_new <- activity_new %>% select(steps, date, interval)

```


- Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
### calculate the total number of steps taken per day
steps_per_day2 <- activity_new %>% 
        group_by(date) %>% 
        do (
                mutate(., sum=sum(steps, na.rm = T)
                       , mean=mean(steps, na.rm = T)
                       , median=median(steps, na.rm = T)
                )
        ) %>%
        distinct(date)

### make a histogram of the total number of steps taken each day
total_days2 <- as.numeric(max(steps_per_day2$date) - min(steps_per_day2$date) + 1)
hist(steps_per_day2$sum
     , main = paste("Total number of steps per day","\n(",total_days2,"-day period)")
     , xlab="Steps per day"
     , ylab="Frequency of days"
)

### calculate and report the mean and median of the total number of steps taken per day
steps_per_day2 %>% select(date, mean, median)

```


- Do these values differ from the estimates from the first part of the assignment? 

Yes.

- What is the impact of imputing missing data on the estimates of the total daily number of steps?

Frequencies seem to be normalized.



## Are there differences in activity patterns between weekdays and weekends?

- Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
nrows <- dim(activity_new)[1]
day_type <- NULL
for (i in 1:nrows){
        date_i <- wday(activity_new[i, "date"])
        if (date_i == 1) {
                day_type <- c(day_type,"weekend")
        }
        else if (date_i == 7) {
                day_type <- c(day_type,"weekend")
        }
        else {
                day_type <- c(day_type,"weekday")
        }
}

activity_new <- data.frame(activity_new,day_type)

```


- Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r}

five_m_int2 <- activity_new %>% 
        group_by(interval, day_type) %>% 
        mutate( mean=mean(steps, na.rm = T)) %>%
        distinct(interval, day_type)

par(mfrow=c(1,2))
with(five_m_int2[five_m_int2$day_type=="weekend", ], 
     plot(interval, mean, type = "l"
          , main=paste("Steps per 5-minute interval","\n(weekend)")
          , xlab="Intervals"
          , ylab="Average number of steps"
     )
)
with(five_m_int2[five_m_int2$day_type=="weekday", ], 
     plot(interval, mean, type = "l"
          , main=paste("Steps per 5-minute interval","\n(weekday)")
          , xlab="Intervals"
          , ylab="Average number of steps"
     )
)

```


